Applying visual inference to a social network model

### 2017-10-12 ASApaper/paper.Rmd
finalize lineups for today, add table to paper of effects.

bibliography: bibliography.bib
output: rticles::asa_article
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, echo = FALSE, message=FALSE,warning=FALSE,
                      out.width = '\\linewidth')
```

@@

```{r senateAll, fig.cap="The four senate collaboration networks that we estimate "}
seobama <- read_csv("../Data/senate/senateobamapres_gsw_25.csv")
to_label <- c("Joseph R. Biden Jr.", "John S. McCain", "Ted Cruz","Marco Rubio",
  "Lindsey O. Graham","Rand Paul", "Bernard Sanders", "Jim Webb", "Mitch McConnell",
  "Harry M. Reid", "Hillary Rodham Clinton", "Amy Jean Klobuchar")
seobama$label <- as.factor(ifelse(seobama$source %in% to_label, seobama$source, ""))
levels(seobama$label) <- c("", "Amy Klobuchar", "Bernie Sanders", "Harry Reid",
                           "Hillary Clinton", "Jim Webb", "John McCain", "Joe Biden",
                           "Lindsey Graham", "Marco Rubio", "Mitch McConnell", 
                           "Rand Paul", "Ted Cruz")
se111clint <- filter(seobama, senate == 111)
se111noclint <- filter(seobama, senate == 111)
se111noclint$target[which(se111noclint$target == "Hillary Rodham Clinton")] <- NA
seobama2 <- seobama %>% filter(senate != 111) %>% bind_rows(se111noclint)

set.seed(56049382)
ggplot(data = seobama2) + 
  geom_net(directed = T, labelon=T, arrowsize = .25, singletons= T, fiteach = T, linewidth = .25, layout.alg = 'fruchtermanreingold', fontsize = 3, repel=T, 
           aes(from_id = source, to_id = target, color = party, label = label),
           size = 1) + 
  theme_net() + 
  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
  theme(legend.position = 'bottom', strip.text = element_text(size = 14)) + 
  facet_wrap(~senate, nrow = 2, labeller = "label_both") + 
  xlim(c(0,1.05)) + 
  ylim(c(0,1.05))
```
```{r senateClinton}
se111clint$Clinton <- 'Yes'
se111noclint$Clinton <- 'No'
clintonSenate <- rbind(se111clint, se111noclint)
clintonSenate$Clinton <- as.factor(clintonSenate$Clinton)
clintonSenate$Clinton <- ordered(clintonSenate$Clinton, levels = c("Yes", "No"))
clintonSenate %>% filter(!(source %in% c("Roland Burris", "Bernard Sanders", "Frank R. Lautenberg", "Judd A. Gregg", "Mary L. Landrieu") & is.na(target))) %>%
ggplot() + 
  geom_net(directed = T, labelon=F, arrowsize = .3, singletons= F, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
           aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au)) + 
  scale_size_continuous(name="# bills\nauthored") +
  theme_net() + 
  scale_shape_manual(values = c(17,16)) + 
  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
  theme(legend.position = 'bottom') + 
  facet_wrap(~Clinton, nrow = 1, labeller = 'label_both')
```

@@

In cosponsorship networks, it is well known that party affiliation and reciprocity of relationships are major influences on structure [@legnet].

@@

After completing the procedure for all model effects, we selected effects to include that converged and had a Wald $p$-value of less than 0.10, and seemed to have a reasonable interpretation for our data.

The most significant effect was the jumping transitive triplet (JTT) parameter for the party covariate, which was estimated to be about -6 with a standard error of 0.11. This effect relies on the number of transitive closures formed between two senators from different parties. Due to the divisive nature of American politics, the party effect is dominant. Another significant effect was the same JTT parameter for the sex covariate, with an estimate of about 3 with a standard error of 0.89. Another significant effect was the covariate-related similarity score-weighted transitive triplets for the number of bills authored by a senator. This effect was estimated at about 10 with standard error of 3.9, and the high positive effect suggests senators tend to collaborate with other senators who author about the same number of bills they do. Finally, we found the same party transitive triplet effect was also significant, with a fitted value of 1.3 and standard error of 0.7.

List of effects: 
\begin{enumerate}
\item jtt - party
\item jtt - sex
\item similiarity trans trip - bills
\item same tt - party
\end{enumerate}

# Simulation Procedure
\begin{enumerate}
\item Fit each model 1,000 times and save the parameter estimates, standard errors, and overall convergence criterion.
\item From the models that converged, compute the mean of the 1,000 estimates of each parameter.
\item Use the mean values from step 2 as the parameter values in the model and simulate 100 observations from the model. These simulated networks will become the "true data."
\item To each simulation, refit the model form from which it was simulated. 
\item Repeat steps 4-5 for various permutations of model forms and parameter values; e.g. double the parameter values, quadruple the parameter values, change the signs on the parameters, etc.
\item Simulate additional "alternative data" from the model in step 3 and the fitted model in step 4. 
\item Place the appropriate node-link diagram of each "true data" observation in a lineup with $M-1$ (M=12? M=9?) "alternative data" node-link diagrams. 
\item Show the lineups to participants, with no participant seeing the same "true data" or "alternative data" plot twice. 
\end{enumerate}

>>>>

preamble: 
  - \usepackage{amssymb}
bibliography: bibliography.bib
output: rticles::asa_article
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, echo = FALSE, message=FALSE,warning=FALSE,
                      out.width = '\\linewidth')
```

@@

```{r senateAll, fig.cap="The four senate collaboration networks that we use as our example data to visually assess the SAOM effects.", fig.height=4}
seobama <- read_csv("../Data/senate/senateobamapres_gsw_25.csv")
to_label <- c("Joseph R. Biden Jr.", "John S. McCain", "Ted Cruz","Marco Rubio",
  "Lindsey O. Graham","Rand Paul", "Bernard Sanders", "Jim Webb", "Mitch McConnell",
  "Harry M. Reid", "Hillary Rodham Clinton", "Amy Jean Klobuchar")
seobama$label <- as.factor(ifelse(seobama$source %in% to_label, seobama$source, ""))
levels(seobama$label) <- c("", "Amy Klobuchar", "Bernie Sanders", "Harry Reid",
                           "Hillary Clinton", "Jim Webb", "John McCain", "Joe Biden",
                           "Lindsey Graham", "Marco Rubio", "Mitch McConnell", 
                           "Rand Paul", "Ted Cruz")
se111clint <- filter(seobama, senate == 111)
se111noclint <- filter(seobama, senate == 111)
se111noclint$target[which(se111noclint$target == "Hillary Rodham Clinton")] <- NA
seobama2 <- seobama %>% filter(senate != 111) %>% bind_rows(se111noclint)

set.seed(56049382)
ggplot(data = seobama2) + 
  geom_net(directed = T, labelon=T, arrowsize = .25, singletons= T, fiteach = T, linewidth = .25, layout.alg = 'fruchtermanreingold', fontsize = 3, repel=T, 
           aes(from_id = source, to_id = target, color = party, label = label),
           size = 1) + 
  theme_net() + 
  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
  theme(legend.position = 'bottom', strip.text = element_text(size = 14)) + 
  facet_wrap(~senate, nrow = 2, labeller = "label_both") + 
  xlim(c(0,1.05)) + 
  ylim(c(0,1.05))
```

```{r senateClinton, fig.cap="We removed Hillary Clinton's ties from the network because she had abnormally high collaboration with senators during the time she was in the 111th senate and before she left office to become Secretary of State.", fig.height=2.5}
se111clint$Clinton <- 'Yes'
se111noclint$Clinton <- 'No'
clintonSenate <- rbind(se111clint, se111noclint)
clintonSenate$Clinton <- as.factor(clintonSenate$Clinton)
clintonSenate$Clinton <- ordered(clintonSenate$Clinton, levels = c("Yes", "No"))
clintonSenate %>% filter(!(source %in% c("Roland Burris", "Bernard Sanders", "Frank R. Lautenberg", "Judd A. Gregg", "Mary L. Landrieu") & is.na(target))) %>%
ggplot() + 
  geom_net(directed = T, labelon=F, arrowsize = .3, singletons= F, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
           aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au)) + 
  scale_size_continuous(name="# bills\nauthored") +
  theme_net() + 
  scale_shape_manual(values = c(17,16)) + 
  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) + 
  theme(legend.position = 'bottom') + 
  facet_wrap(~Clinton, nrow = 1, labeller = 'label_both')
```

@@

In legislative cosponsorship networks, it is well known that party affiliation and reciprocity of relationships are major influences on structure [@legnet].

@@

After completing the procedure for all model effects, we selected effects whose estimates converged, had a Wald $p$-value of less than 0.10, and seemed to have a reasonable interpretation for our data according to well-known properties of legislative networks [@legnet].

The most significant effect was the jumping transitive triplet (JTT) parameter for the party covariate, which was estimated to be about -6 with a standard error of 0.11. This effect relies on the number of transitive closures formed between two senators from different parties. The negative estimate is an indication that forming transitive ties between two people from different parties is discouraged, which tracks with the divisive nature of American politics, where party affilitation is dominant. Another significant effect was the same JTT parameter for the sex covariate, with an estimate of about 3 with a standard error of 0.89. Another significant effect was the covariate-related similarity score-weighted transitive triplets for the number of bills authored by a senator. This effect was estimated at about 10 with standard error of 3.9, and the high positive effect suggests senators tend to collaborate with other senators who author about the same number of bills they do. Finally, we found the same party transitive triplet effect was also significant, with a fitted value of 1.3 and standard error of 0.7.

```{r geteffects, eval = FALSE}
# used for table below
initeff <- read_csv("../Data/senate/sigEffsSenate912.csv")
initeff %>% filter(shortName == "sameX")
#density & -- & $\sum_j x_{ij}$ & & & \\
#reciprocity & -- & $\sum_j x_{ij}x_{ji}$ & & & \\
```

List of effects: 
\begin{table}
\scalebox{0.9}{
\begin{tabular}{p{2cm}|p{2.1cm}|c|l|p{1.75cm}|p{2cm}}
{\bf Effect name} & {\bf Interaction Variable} & {\bf Formula} & {\bf Picture} & {\bf Initial estimate} & {\bf Wald $p$-value} \\
\hline 
jumping transitive triplet & party & $\sum_{j\neq h} x_{ij}x_{ih}x_{hj}\cdot \mathbb{I}(p_i = p_h \neq p_j)$ & \includegraphics[width=.6in]{img/jttp.png} & -5.884 & $<0.0001$\\
jumping transitive triplet & sex & $\sum_{j\neq h} x_{ij}x_{ih}x_{hj}\cdot \mathbb{I}(s_i = s_h \neq s_j)$ & \includegraphics[width=.6in]{img/jtts.png}  & 3.335 & 0.0002 \\
similarity transitive triplet & bills & $\sum_{j} x_{ij}x_{ih}x_{hj}\cdot (sim^b_{ij} - \overline{sim}^b)^*$ & \includegraphics[width=1in]{img/simttb.png} & 9.821 & 0.0128 \\
same transtive triplet & party & $\sum_{j} x_{ij}x_{ih}x_{hj}\cdot \mathbb{I}(p_i = p_j)$ & \includegraphics[width=1in]{img/samettp.png}  & 1.306 & 0.0642 \\
same & party & $\sum_j x_{ij}\mathbb{I}(p_i = p_j)$ & \includegraphics[width=.6in]{img/samep.png}  & 0.363 & 0.0074
\end{tabular}
}
\caption{\label{tab:effects} The additional effects we used in the SAOMs fit to the senate data. * - $sim^b_{ij} = \frac{\max_{hk}|b_h - b_k| - |b_i - b_j|}{\max_{hk}|b_h - b_k|}$ is the similarity score between two senators based on the number of bills authored, and $\overline{sim}^b = \frac{1}{n(n-1)}\sum_{i\neq j} sim^b_{ij}$ is the average bill similarity score between any two senators.}
\end{table}
# Simulation Procedure
\begin{enumerate}
\item Fit each model 1,000 times and save the parameter estimates, standard errors, and overall convergence criterion.
\item From the models that converged, compute the mean of the 1,000 estimates of each parameter.
\item Use the mean values from step 2 as the parameter values in the model and simulate 100 observations from the model. These simulated networks will become the "true data."
\item To each simulation, refit the model form from which it was simulated. 
\item Repeat steps 4-5 for various permutations of model forms and parameter values; e.g. double the parameter values, quadruple the parameter values, change the signs on the parameters, etc.
\item Simulate additional "alternative data" from the model in step 3 and the fitted model in step 4. 
\item Place the appropriate node-link diagram of each "true data" observation in a lineup with $M-1$ (M=12? M=9?) "alternative data" node-link diagrams. 
\item Show the lineups to participants, with no participant seeing the same "true data" or "alternative data" plot twice. 
\end{enumerate}

### 2017-10-12 GG-10-12/slides.Rmd
finalize lineups for today, add table to paper of effects.

@@

```{r trialE}
dat <- read_csv("LineupData/basicx-5sd346231pickalt.csv")
ggplot(data = dat) + 
        geom_net(aes(from_id = from, to_id = to), 
                 arrow = arrow(type = 'open', length = unit(2, "points") ), 
                 linewidth = .25, singletons = T, fiteach = T, directed = T, 
                 color = 'black', arrowgap = .015, arrowsize = .3, size = .5) + 
        theme_net() + 
        theme(panel.background = element_rect(color = 'black')) +
        facet_wrap(~ord)
```

@@

```{r trialG}
dat <- read_csv("LineupData/samettpx2sd23452pickalt.csv")
ggplot(data = dat) + 
        geom_net(aes(from_id = from, to_id = to), 
                 arrow = arrow(type = 'open', length = unit(2, "points") ), 
                 linewidth = .25, singletons = T, fiteach = T, directed = T, 
                 color = 'black', arrowgap = .015, arrowsize = .3, size = .5, 
                 layout.alg = "fruchtermanreingold") + 
        theme_net() + 
        theme(panel.background = element_rect(color = 'black')) +
        facet_wrap(~ord)
```

@@

```{r trialK}
dat <- read_csv("LineupData/samettpx-1sd35792pickalt.csv")
ggplot(data = dat) + 
        geom_net(aes(from_id = from, to_id = to), 
                 arrow = arrow(type = 'open', length = unit(2, "points") ), 
                 linewidth = .25, singletons = T, fiteach = T, directed = T, 
                 color = 'black', arrowgap = .015, arrowsize = .3, size = .5) + 
        theme_net() + 
        theme(panel.background = element_rect(color = 'black')) +
        facet_wrap(~ord)
```

>>>>

## Trial D2

```{r trialD2}
#same as K with KK instead of FR layout
dat <- read_csv("LineupData/samettpx-1sd35792pickalt.csv")
ggplot(data = dat) + 
        geom_net(aes(from_id = from, to_id = to), 
                 arrow = arrow(type = 'open', length = unit(2, "points") ), 
                 linewidth = .25, singletons = T, fiteach = T, directed = T, 
                 color = 'black', arrowgap = .015, arrowsize = .3, size = .5) + 
        theme_net() + 
        theme(panel.background = element_rect(color = 'black')) +
        facet_wrap(~ord)
```

@@

```{r trialE}
dat <- read_csv("LineupData/basicx-5sd346231pickalt.csv")
ggplot(data = dat) + 
        geom_net(aes(from_id = from, to_id = to), 
                 arrow = arrow(type = 'open', length = unit(2, "points") ), 
                 linewidth = .25, singletons = T, fiteach = T, directed = T, 
                 color = 'black', arrowgap = .015, arrowsize = .3, size = .5,
                 layout.alg = 'fruchtermanreingold') + 
        theme_net() + 
        theme(panel.background = element_rect(color = 'black')) +
        facet_wrap(~ord)
```

@@

```{r trialG}
dat <- read_csv("LineupData/samettpx2sd23452pickalt.csv")
ggplot(data = dat) + 
        geom_net(aes(from_id = from, to_id = to), 
                 arrow = arrow(type = 'open', length = unit(2, "points") ), 
                 linewidth = .25, singletons = T, fiteach = T, directed = T, 
                 color = 'black', arrowgap = .015, arrowsize = .3, size = .5) + 
        theme_net() + 
        theme(panel.background = element_rect(color = 'black')) +
        facet_wrap(~ord)
```

@@

```{r trialK}
dat <- read_csv("LineupData/samettpx-1sd35792pickalt.csv")
ggplot(data = dat) + 
        geom_net(aes(from_id = from, to_id = to), 
                 arrow = arrow(type = 'open', length = unit(2, "points") ), 
                 linewidth = .25, singletons = T, fiteach = T, directed = T, 
                 color = 'black', arrowgap = .015, arrowsize = .3, size = .5, 
                 layout.alg = 'fruchtermanreingold') + 
        theme_net() + 
        theme(panel.background = element_rect(color = 'black')) +
        facet_wrap(~ord)
```
